Combining checkpointing and data compression\\ for high frequency seismic imaging

### 2018-10-03 paper/paper.tex
some language changes

\section{Compression algorithms}
\subsection{Lossless}
For our initial experiments, we used the python package \emph{blosc} \cite{blosc}. It includes implementations for
6 different lossless compression algorithms, namely ZLIB, ZSTD, BLOSCLZ,
LZ4, LZ4HC and Snappy. We did a parameter sweep over all available parameters to find the best-performing combinations. 
The best compression ratio we saw in this study was 1.18x which wasn't very promising.
\subsection{Lossy}
Next we tried lossy compression using ZFP \cite{lindstrom2014fixed}. To use ZFP from python, 
we first wrote a python wrapper over the reference implementation of ZFP, which is in C. We plan to release this wrapper
in the open source community shortly.

@@

We tried all three modes of ZFP, namely fixed-tolerance, fixed-precision and fixed-rate modes. Figure \ref{fig:tolerance_cf_plot}
shows the compression ratios we saw for a variety of compression settings for the three modes. As expected, the highest compression efficiency was
observed in the fixed-tolerance mode albeit with highly unpredictable compression ratios. Figure \ref{fig:decompressed_error} shows the
spatial distribution of the error introduced after compression and decompression using fixed-tolerance mode. Since, in our experiments, the time
taken to compress a checkpoint was in the same order of magnitude as the time taken to compute a timestep, we build a
performance model in the next section to evaluate the utility of compression in different scenarios of problem setups and test
platforms.

@@

\section{Performance model for a combination of Revolve and compression}
\label{sec:performance_model}
We work with the assumption that the computation of a single forward time step takes the same wall time
as the computation of a single reverse time step - calling this $\mathbf{C}$. If the size of a single 
timestep in memory is $\mathbf{S}$ and the simulation involves $\mathbf{N}$ timesteps, the minimum
wall time required to run a single forward-adjoint evaluation is given by:
\begin{equation}
T_N = 2 \cdot \mathbf{C} \cdot \mathbf{N}
\end{equation}
However, this would require $\mathbf{S} \cdot \mathbf{N}$ memory. On a hardware platform where enough memory
is available, this would be the optimal strategy to compute the adjoint solution.

When the memory available on a system is less than $\mathbf{S} \cdot \mathbf{N}$, Revolve provides
an optimal strategy to solve for the adjoint field by storing a subset of the $\mathbf{N}$ total checkpoints
and recomputing the remaining ones. The overhead introduced by this method can be broken down into
the recomputation overhead $\mathbf{O}_R$ and the storage overhead $\mathbf{O}_S$. The recomputation
overhead is simply the amount of time spent in recomputation, given by:
\begin{equation}
\mathbf{O}_R(N, M) = p(N, M) \cdot \mathbf{C}
\end{equation}
where $p(N, M)$ is the minimum number of recomputed steps from \cite{griewank2000algorithm}, reproduced
here in equation \ref{eqn:recompute}.
\begin{figure*}
\begin{equation}
p(N, M) = \begin{cases}
      N(N-1) /2, & \text{if}\ M=1 \\
      \min\limits_{1<=\widetilde{N}<=N} \{\widetilde{N} + \mathbf{O}_R(\widetilde{N}, M) + \mathbf{O}_R(N-\widetilde{N}, M-1)\}, & \text{if}\ M>1
    \end{cases}
    \label{eqn:recompute}
\end{equation}
\end{figure*}
In equation \ref{eqn:recompute}, M is the number of checkpoints that can be stored in memory. Note that for $M >=N$, $\mathbf{O}_R$
would be zero. For $M < N$, $\mathbf{O}_R$ goes up rather quickly as M is reduced in relation to N.

In a perfect implementation, the storage overhead $\mathbf{O}_S$ might be zero, since the computation could
be done "in-place", but in practice, checkpoints are generally stored in a separate section of memory and they
need to be transferred to a "computational" section of the memory where the compute is performed, and then
the results copied back to the checkpointing memory. This copying is a common feature of checkpointing
implementations, and might pose a non-trivial overhead in scenarios where the computation is heavily bandwidth-bound. 
This storage overhead is given by:
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot \frac{\mathbf{S}}{\mathbf{B}} + \mathbf{N}_R \cdot \frac{\mathbf{S}}{\mathbf{B}}
\label{eqn:storage}
\end{equation}
where $\mathbf{N}_W$ is the total number of times Revolve writes checkpoints for a single run, $ \mathbf{N}_R$ 
is the number of times checkpoints are read, and $\mathbf{B}$ is the memory bandwidth of the target system.

The total time to solution in this scenario becomes:
\begin{equation}
T_R = 2 \cdot \mathbf{C} \cdot \mathbf{N} + \mathbf{O}_R(N, M) + \mathbf{O}_S
\end{equation}

Adding compression to the mix will increase the number of checkpoints available ($M$ in equation \ref{eqn:recompute}),
hence reducing $\mathbf{O}_R$ while adding overheads related to compression and decompression in $\mathbf{O}_S$,
as well as some error introduced by the compression. Assuming that the error remains within the allowed tolerance,
we hypothesize that, at least for some circumstances, the reduction in $\mathbf{O}_R$ should offset the increase in 
$\mathbf{O}_S$, i.e. lead to an overall decrease in the total time to solution ($T$). To build a performance model that
enables us to find the regions where compression pays off, we assume that the compression algorithm behaves uniformly
across the different time steps of the simulation, i.e. that we get the same compression ratio, compression time and 
decompression time, no matter which of the $N$ possible checkpoints we try to compress/decompress. The storage overhead
now becomes:
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} + t_c) + \mathbf{N}_R \cdot (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} + t_d)
\end{equation}
where $\mathbf{F}$ is the compression ratio (i.e. the ratio between the uncompressed and compressed checkpoint), and $t_c$
and $t_d$ are compression and decompression times, respectively. At the same time, the recomputation overhead goes down
because $\mathbf{F}$ times more checkpoints are now available.

@@

TThis work was supported by the U.S. Department of Energy, Office of Science,
Office of Advanced Scientific Computing Research, Applied Mathematics and
Computer Science programs under contract number DE-AC02-06CH11357.
We would also like to acknowledge the support from the SINBAD II project and
the member organizations of the SINBAD Consortium.

>>>>

\section{Compression algorithms}
\subsection{Lossless}
We use the python package \emph{blosc}~\cite{blosc}, which includes implementations for
six different lossless compression algorithms, namely ZLIB, ZSTD, BLOSCLZ,
LZ4, LZ4HC and Snappy. An exhaustive search over all available parameters 
resulted in a maximum compression ratio of 1.18x. We therefore focus on lossy
compression in the remainder of this work.

\subsection{Lossy}
We use the lossy compression package ZFP~\cite{lindstrom2014fixed} developed in
C. To use ZFP from python, we developed a python wrapper for the reference
implementation of ZFP.

@@

ZFP supports three compression modes, namely fixed-tolerance, fixed-precision
and fixed-rate. The fixed-tolerance mode limits the absolute error, while the
fixed-precision mode limits the relative error to some user-specified value.
The fixed-rate mode achieves a guaranteed compression ratio requested by the
user, but does not provide any bounds on accuracy loss.
Figure~\ref{fig:tolerance_cf_plot} shows compression ratios for a variety of
settings for the three modes.

The highest compression efficiency TODO define this is achieved in the
fixed-tolerance mode albeit with highly unpredictable compression ratios. Figure
\ref{fig:decompressed_error} shows the spatial distribution of the error after
compression and decompression, compared to the original field, using
fixed-tolerance mode.

In our experiments using Devito, the time required to compress a checkpoint was
in the same orderof magnitude as the time taken to compute a timestep.

@@

\section{Performance model for a combination of Revolve and compression}
\label{sec:performance_model}
We assume that the computation of a single forward time step takes the same wall
time as the computation of a single reverse time step. We denote this time as
$\mathbf{C}$. For a simulation with $\mathbf{N}$ timesteps, the minimum wall time required
for the full forward-adjoint evaluation is given by
\begin{equation}
T_N = 2 \cdot \mathbf{C} \cdot \mathbf{N}.
\end{equation}
If the size of a single timestep in memory is given by $\mathbf{S}$, this
requires a memory of at least size $\mathbf{S} \cdot \mathbf{N}$. If sufficient memory
is available, no checkpointing or compression is needed.

If the memory is smaller than $\mathbf{S} \cdot \mathbf{N}$, Revolve provides
a strategy to solve for the adjoint field by storing a subset of the $\mathbf{N}$ total checkpoints
and recomputing the remaining ones. The overhead introduced by this method can be broken down into
the recomputation overhead $\mathbf{O}_R$ and the storage overhead $\mathbf{O}_S$. The recomputation
overhead is the amount of time spent in recomputation, given by
\begin{equation}
\mathbf{O}_R(N, M) = p(N, M) \cdot \mathbf{C},
\end{equation}
where $p(N, M)$ is the minimum number of recomputed steps from \cite{griewank2000algorithm}, reproduced
here in equation \ref{eqn:recompute}.
\begin{figure*}
\begin{equation}
p(N, M) = \begin{cases}
      N(N-1) /2, & \text{if}\ M=1 \\
      \min\limits_{1<=\widetilde{N}<=N} \{\widetilde{N} + \mathbf{O}_R(\widetilde{N}, M) + \mathbf{O}_R(N-\widetilde{N}, M-1)\}, & \text{if}\ M>1
    \end{cases}
    \label{eqn:recompute}
\end{equation}
\end{figure*}
In equation \ref{eqn:recompute}, M is the number of checkpoints that can be
stored in memory. Note that for $M >=N$, $\mathbf{O}_R$ would be zero. For $M <
N$, $\mathbf{O}_R$ grows rapidly as M is reduced relative to N.

In an ideal implementation, the storage overhead $\mathbf{O}_S$ might be zero, since the computation could
be done "in-place", but in practice, checkpoints are generally stored in a separate section of memory and they
need to be transferred to a "computational" section of the memory where the compute is performed, and then
the results copied back to the checkpointing memory. This copying is a common feature of checkpointing
implementations, and might pose a non-trivial overhead in scenarios where the computation is heavily bandwidth-bound. 
This storage overhead is given by
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot \frac{\mathbf{S}}{\mathbf{B}} +
\mathbf{N}_R \cdot \frac{\mathbf{S}}{\mathbf{B}}.
\label{eqn:storage}
\end{equation}
where $\mathbf{N}_W$ is the total number of times Revolve writes checkpoints for a single run, $ \mathbf{N}_R$ 
is the number of times checkpoints are read, and $\mathbf{B}$ is the memory bandwidth of the target system.

The total time to solution in this scenario becomes
\begin{equation}
T_R = 2 \cdot \mathbf{C} \cdot \mathbf{N} + \mathbf{O}_R(N, M) + \mathbf{O}_S.
\end{equation}

By using compression, the size of each checkpoint is reduced and therefore the
number of checkpoints available is increased ($M$ in equation
\ref{eqn:recompute}). This reduces the recomputation overhead $\mathbf{O}_R$,
while at the same time adding overheads related to compression and decompression
in $\mathbf{O}_S$.
To be beneficial, the reduction in $\mathbf{O}_R$ must offset the increase in 
$\mathbf{O}_S$, leading to an overall decrease in the time to solution $T$.

Our performance model assumes
 that the compression algorithm behaves uniformly
across the different time steps of the simulation, i.e. that we get the same compression ratio, compression time and 
decompression time, no matter which of the $N$ possible checkpoints we try to compress/decompress. The storage overhead
now becomes
\begin{equation}
\mathbf{O}_{SR} = \mathbf{N}_W \cdot (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} +
t_c) + \mathbf{N}_R \cdot (\frac{\mathbf{S}}{\mathbf{F}\mathbf{B}} + t_d),
\end{equation}
where $\mathbf{F}$ is the compression ratio (i.e. the ratio between the uncompressed and compressed checkpoint), and $t_c$
and $t_d$ are compression and decompression times, respectively. At the same
time, the recomputation overhead decreases
because $\mathbf{F}$ times more checkpoints are now available.

@@

This work was supported by the U.S. Department of Energy, Office of Science,
Office of Advanced Scientific Computing Research, Applied Mathematics and
Computer Science programs under contract number DE-AC02-06CH11357.
We would also like to acknowledge the support from the SINBAD II project and
the member organizations of the SINBAD Consortium.
